Lay Perceptions of Crowd-Scientific Findings: The Risks of Variability and Lack of Consensus

### 2021-10-07 manuscript/manuscript.Rmd
edited manuscript intro to crowd science, edited talk (title plot, outline, intro to crowd science)

keywords          : "Crowd science, Many analysts, Multi-analyst, Variability, Research credibility"
wordcount         : "3,627"

@@

The credibility of scientific research is in doubt, among lay consumer [@hornsey2017] and scientist [@pashler2012] alike. Several tools have been proposed to combat this "crisis of confidence" (Ibid., p. 528). One such tool is the crowd science approach, which leverages a large number of individuals or teams at specific stages of the research process [@uhlmann2019]. Does crowd science meet its promise -- to improve the credibility of scientific research -- in reality?  
We explore the effects of scientific findings emerging from a crowd of researchers (vs. a typical research collaboration) on lay consumers' posterior beliefs, perceptions of credibility, confidence in an aggregate effect size estimate, and ratings of bias and error. We focus on crowdsourced data analysis, also known as the 'many analysts' or 'multi-analyst' approach: giving the same dataset to different teams of scientists, who independently analyze it to answer the same research question and/or estimate a parameter of interest. We compare the effects of providing (a) a single, aggregate parameter estimate (the single-analyst condition) vs. multiple parameter estimates that (b) vary slightly and are all positive, leading to the same qualitative conclusion (the "multi-consistent" condition) or (c) vary widely and are of both signs, leading to differing qualitative conclusions (the "multi-inconsistent" condition).    
According to previous work, crowdsourced data analysis should increase the credibility of scientific findings; Table 1 provides an overview of the proposed confidence- and credibility-related benefits of the multi-analyst approach. Normatively, we expect laypeople who observe several scientists independently come to the same qualitative conclusions -- compared to a single research team that comes to a single, joint conclusion -- to be more swayed by the findings, as it seems to reflect consensus between various approaches. After all, it's a hallmark of rigorous research to triangulate various methodological approaches and provide converging evidence [@jick1979; @turner2017], and to provide a "sensitivity analysis" or "robustness checks" to examine whether and how the findings change as a result of alternative analytic specifications [@munoz2018; @steegen2016]. As argued in Aczel et al. [-@aczel2020, p. 562], "The main argument for the importance of performing robustness checks over reasonable variations in modelling choices is to increase confidence in the obtained results: ideally, results should be reasonably unaffected by a researcher's idiosyncratic choice (…) when reasonable alternatives exist." The same argument has been made for triangulation, which, according to Jick [-@jick1979, p. 608], "allows researchers to be more confident of their results." We would similarly expect readers of multi-analyst reports that yield consistent findings to be more confident in the results. Although lay observers, in such cases, may be less certain that the precise aggregate parameter estimate is exactly correct, they should be more likely to positively update their beliefs about the reported phenomenon, and less likely to assume that the overall estimate stems from bias or error. Thus, when the results generated by independent analysts are largely consistent, we expect an increase in the sway of scientific findings. However, when laypeople observe several scientists independently come to differing qualitative conclusions, we expect the multi-analyst method to backfire; when results across many analysts vary widely and lack consensus in their qualitative conclusions (which, arguably, often reflects the reality of large-scale science collaborations), we expect a decrease in the sway of scientific findings.  

>>>>

keywords          : "Crowd science, Many analysts, Multi-analyst, Variability, Research credibility"
wordcount         : "3,710"

@@

The credibility of scientific research is in doubt, among lay consumer [@hornsey2017] and scientist [@pashler2012] alike. Several tools have been proposed to combat this "crisis of confidence" (Ibid., p. 528). One such tool is the crowd science approach, which leverages a large number of individuals or teams at specific stages of the research process [@uhlmann2019].  
Crowd science is defined as "the organization of scientific research in open collaborative projects"  [@franzoni2014, p. 1]. Crowd-scientific projects vary in their degree of selectivity: some are open to the general public (such projects are often referred to as "citizen science"), while others require the expertise of specialists [e.g., scientists\; @uhlmann2019, p. 714]. Here, we focus on crowdsourced data analysis, also known as the 'many analysts' or 'multi-analyst' approach: giving the same dataset to different teams of scientists, who independently analyze it to answer the same research question and/or estimate a parameter of interest.  
We explore the effects of scientific findings emerging from a crowd of researchers (vs. a typical research collaboration) on lay consumers' posterior beliefs, perceptions of credibility, confidence in an aggregate effect size estimate, and ratings of bias and error. We compare the effects of providing (a) a single, aggregate parameter estimate (the single-analyst condition) vs. multiple parameter estimates that (b) vary slightly and are all positive, leading to the same qualitative conclusion (the "multi-consistent" condition) or (c) vary widely and are of both signs, leading to differing qualitative conclusions (the "multi-inconsistent" condition).    
According to previous work, crowdsourced data analysis should increase the credibility of scientific findings; Table 1 provides an overview of the proposed confidence- and credibility-related benefits of the multi-analyst approach. Does crowd science meet its promise -- to improve the credibility of scientific research -- in reality? Normatively, we expect laypeople who observe several scientists independently come to the same qualitative conclusions -- compared to a single research team that comes to a single, joint conclusion -- to be more swayed by the findings, as it seems to reflect consensus between various approaches. After all, it's a hallmark of rigorous research to triangulate various methodological approaches and provide converging evidence [@jick1979; @turner2017], and to provide a "sensitivity analysis" or "robustness checks" to examine whether and how the findings change as a result of alternative analytic specifications [@munoz2018; @steegen2016]. As argued in Aczel et al. [-@aczel2020, p. 562], "The main argument for the importance of performing robustness checks over reasonable variations in modelling choices is to increase confidence in the obtained results: ideally, results should be reasonably unaffected by a researcher's idiosyncratic choice (…) when reasonable alternatives exist." The same argument has been made for triangulation, which, according to Jick [-@jick1979, p. 608], "allows researchers to be more confident of their results." We would similarly expect readers of multi-analyst reports that yield consistent findings to be more confident in the results. Although lay observers, in such cases, may be less certain that the precise aggregate parameter estimate is exactly correct, they should be more likely to positively update their beliefs about the reported phenomenon, and less likely to assume that the overall estimate stems from bias or error. Thus, when the results generated by independent analysts are largely consistent, we expect an increase in the sway of scientific findings. However, when laypeople observe several scientists independently come to differing qualitative conclusions, we expect the multi-analyst method to backfire; when results across many analysts vary widely and lack consensus in their qualitative conclusions (which, arguably, often reflects the reality of large-scale science collaborations), we expect a decrease in the sway of scientific findings.  

### 2021-10-07 talk/index.Rmd
edited manuscript intro to crowd science, edited talk (title plot, outline, intro to crowd science)

estimates <- data.frame(
  Estimate = sample(
    x = seq(-0.3, 0.3, .001), 
    size = 100
    ),
  se = sample(
    x = seq(0.006, 0.05, .001),
    replace = TRUE,
    size = 100
  )) %>% 
  arrange(Estimate) %>% 
  mutate(Team = 1:100,
         Conclusion = case_when(
           Estimate - 1.96 * se < 0 & Estimate + 1.96 * se < 0 ~ "Negative Evidence",
           Estimate - 1.96 * se < 0 & Estimate + 1.96 * se > 0 ~ "Inconclusive Evidence",
           Estimate - 1.96 * se > 0 & Estimate + 1.96 * se > 0 ~ "Positive Evidence"
         )) %>% 
  mutate(Conclusion = factor(Conclusion,
                             levels = c("Negative Evidence",
                                        "Inconclusive Evidence",
                                        "Positive Evidence")))

title_plot <- ggplot(data = estimates, 
                     mapping = aes(x = Team, y = Estimate, color = Conclusion)) +
  geom_point(size = 2) +
  geom_errorbar(mapping = aes(ymin = Estimate - 1.96 * se, 
                              ymax = Estimate + 1.96 * se),
                width = 3) + 
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(y = "Standardized Effect Size") +
  scale_color_discrete(type = c("red", "orange", "green4")) +
  theme(legend.position = c(0.3, 0.8),
        text = element_text(size = 19))

@@

.pull-right[
[`r icon_style(icons::fontawesome("twitter"), scale = 2, fill = "#035AA6")` $~~$ twitter.com/shilaan01](https://twitter.com/shilaan01)    
[`r icon_style(icons::fontawesome("globe"), scale = 2, fill = "#035AA6")` $~~$ shilaan.rbind.io](https://shilaan.rbind.io)     
[`r icon_style(icons::fontawesome("github"), scale = 2, fill = "#035AA6")` $~~$ github.com/shilaan](https://github.com/shilaan)  
[`r icon_style(icons::fontawesome("paper-plane"), scale = 2, fill = "#035AA6")` $~~$ shilaan@stanford.edu](mailto:shilaan@stanford.edu)

@@

@@

---
# Crowdsourced Data Analysis

>>>>

estimates <- data.frame(
  Estimate = sample(
    x = c(seq(-0.3, 0.3, .001), seq(-0.1, 0.1, .0001)), 
    size = 100
    ),
  se = sample(
    x = seq(0.003, 0.04, .001),
    replace = TRUE,
    size = 100
  )) %>% 
  arrange(Estimate) %>% 
  mutate(Team = 1:100,
         Conclusion = case_when(
           Estimate - 1.96 * se < 0 & Estimate + 1.96 * se < 0 ~ "Negative Evidence",
           Estimate - 1.96 * se < 0 & Estimate + 1.96 * se > 0 ~ "Inconclusive Evidence",
           Estimate - 1.96 * se > 0 & Estimate + 1.96 * se > 0 ~ "Positive Evidence"
         )) %>% 
  mutate(Conclusion = factor(Conclusion,
                             levels = c("Negative Evidence",
                                        "Inconclusive Evidence",
                                        "Positive Evidence")))

title_plot <- ggplot(data = estimates, 
                     mapping = aes(x = Team, y = Estimate, color = Conclusion)) +
  geom_point(size = 2) +
  geom_errorbar(mapping = aes(ymin = Estimate - 1.96 * se, 
                              ymax = Estimate + 1.96 * se),
                width = 3) + 
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(x = "Science Team", 
       y = "Obtained Effect Size",
       color = "Conclusion of Science Team") +
  scale_color_discrete(type = c("red", "orange", "green4")) +
  theme(legend.position = c(0.35, 0.8),
        text = element_text(size = 19))

@@

.pull-right[
$~~$

[`r icon_style(icons::fontawesome("twitter"), scale = 2, fill = "#035AA6")` $~~$ twitter.com/shilaan01](https://twitter.com/shilaan01)    
[`r icon_style(icons::fontawesome("globe"), scale = 2, fill = "#035AA6")` $~~$ shilaan.rbind.io](https://shilaan.rbind.io)     
[`r icon_style(icons::fontawesome("github"), scale = 2, fill = "#035AA6")` $~~$ github.com/shilaan](https://github.com/shilaan)  
[`r icon_style(icons::fontawesome("paper-plane"), scale = 2, fill = "#035AA6")` $~~$ shilaan@stanford.edu](mailto:shilaan@stanford.edu)

@@

`r icon_style(icons::fontawesome("users"), scale = 2, fill = "#035AA6")` $~~$ **Crowdsourcing Science** 
`r icon_style(icons::fontawesome("chart-bar"), scale = 2, fill = "#035AA6")` $~~~$ **Crowdsourced Data Analysis**  
`r icon_style(icons::fontawesome("question-circle"), scale = 2, fill = "#035AA6")` $~~~$ **Lay Perceptions: Are Crowd-Scientific Findings More Credible?**   
`r icon_style(icons::fontawesome("tools"), scale = 2, fill = "#035AA6")` $~~~$ **Method**
`r icon_style(icons::fontawesome("chart-line"), scale = 2, fill = "#035AA6")` $~~~$ **Results**
`r icon_style(icons::fontawesome("comment"), scale = 2, fill = "#035AA6")` $~~~$ **Discussion**

---
.center[
# Crowd Science
]

### The organization of scientific research in open and collaborative projects

Highly **inclusive** (sourcing citizens) or highly **selective** (sourcing specialists)

Used at different **stages** of the research process

.pull-left[

.pull-left[

`r icon_style(icons::fontawesome("coins"), scale = 1, fill = "#035AA6")` $~~~$ **Funding**
`r icon_style(icons::fontawesome("lightbulb"), scale = 1, fill = "#035AA6")`  $~~~~$ **Ideation**
`r icon_style(icons::fontawesome("project-diagram"), scale = 1, fill = "#035AA6")`  $~~$  **Study design**
`r icon_style(icons::fontawesome("folder-open"), scale = 1, fill = "#035AA6")`  $~~$  **Data collection**

]

.pull-right[

`r icon_style(icons::fontawesome("laptop-code"), scale = 1, fill = "#035AA6")`  $~~$   **Data analysis**
`r icon_style(icons::fontawesome("feather-alt"), scale = 1, fill = "#035AA6")`  $~~~$    **Writing**
`r icon_style(icons::fontawesome("file-pdf"), scale = 1, fill = "#035AA6")`  $~~~$    **Peer review**
`r icon_style(icons::fontawesome("check-double"), scale = 1, fill = "#035AA6")`  $~~$   **Replication**

]

]

.pull-right[

]

---
background-image: url("images/meteor.jpeg")
background-position: 90% 73%
background-size: 42%

.center[
# Crowd Science
]

### The organization of scientific research in open and collaborative projects

Highly **inclusive** (sourcing citizens) or highly **selective** (sourcing specialists)

Used at different **stages** of the research process

.pull-left[

.pull-left[

`r icon_style(icons::fontawesome("coins"), scale = 1, fill = "#035AA6")` $~~~$ **Funding**
`r icon_style(icons::fontawesome("lightbulb"), scale = 1, fill = "#035AA6")`  $~~~~$ **Ideation**
`r icon_style(icons::fontawesome("project-diagram"), scale = 1, fill = "#035AA6")`  $~~$  **Study design**
`r icon_style(icons::fontawesome("folder-open"), scale = 1, fill = "#035AA6")`  $~~$  **Data collection**

]

.pull-right[

`r icon_style(icons::fontawesome("laptop-code"), scale = 1, fill = "#035AA6")`  $~~$   **Data analysis**
`r icon_style(icons::fontawesome("feather-alt"), scale = 1, fill = "#035AA6")`  $~~~$    **Writing**
`r icon_style(icons::fontawesome("file-pdf"), scale = 1, fill = "#035AA6")`  $~~~$    **Peer review**
`r icon_style(icons::fontawesome("check-double"), scale = 1, fill = "#035AA6")`  $~~$   **Replication**

]

]

.pull-right[

]

.right[

$~~$  
$~~$  
The great meteor storm of 1833: birth of *Citizen Science*

]
---
.center[
# Crowdsourced data analysis

]

### Different teams of scientists independently analyze the same dataset

Also known as the **many-analysts** or **multi-analyst** approach

@@
